---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# PooledPrevalence

<!-- badges: start -->
<!-- badges: end -->

A set of tools to help design a laboratory based prevalence study harnessing the pooling of laboratory samples to increase resource efficiency.
The tools provide guidance in choosing the number of samples to pool and the total number of tests to perform, focusing on achieving the maximum estimation performance (low uncertainty and estimation error) while keeping the total amount of tests low.
In the package there are functions to simulate study outcomes, optimize the study design, and finally, analyze the results of a pooled study to retrieve the prevalence estimates.
The package is based on a Bayesian framework, with the possibility to perform a full hierarchical analysis including uncertainty due to the sensitivity in acquiring the testing material.
The package was developed at the occasion of the COVID-19 pandemic, but can be utilized in every laboratory based prevalence study.

This package is produced as a companion to the following guidance from the European Center for Disease Prevention and Control: https://www.ecdc.europa.eu/sites/default/files/documents/Methodology-estimating-point-prevalence%20-SARS-CoV-2-infection-pooled-RT-PCR-testing.pdf.

## Installation

You can install the released version of PooledPrevalence with:

``` r
# install.packages("devtools")
devtools::install_github("EU-ECDC/PooledPrevalence")
```
## Basic usage

The main goal of the package is to estimate prevalence in a population/risk group, given the results of a pooled laboratory test. Let's assume a test with the following design: 2000 individual biological samples divided in 10 samples per testing pool (pool size), for a total of 200 pools. In this experiment we observe 30 positive pools (15%). With `get_estimates()` it is possible to estimate the underlying prevalence:

```{r base_estimation}
library(PooledPrevalence)

results <- get_estimates(10, 200, 30)

print(results$estimates)
```

The object `results` contains both the estimates and, if a Bayesian method is used, the posterior samples of the distribution. The samples allows the inspection of the full posterior distribution of the prevalence.

```{r plot_posterior, dpi=400}

results <- get_estimates(10, 200, 30)

plot(density(results$samples))
```

With Hierarchical Bayesian (HB) estimation is possible to account for the risk of false negatives due to incorrect sampling of material. Notice that the resulting estimates are therefore slightly higher. HB allows also a more correct evaluation of the uncertainty and should be the preferred method for the final analyses.

```{r Hierarchical_Bayes, dpi=400}

results <- get_estimates(10, 200, 30, method = 'HB')

plot(density(results$samples))
```

## Hypothesis testing

Having access to the posterior samples allows Bayesian hypothesis testing for decision making. For example, a public health authority may want to remove in a geographical region a certain control measure only if the average prevalence is below 2% with a > 90% probability:

```{r hypothesis_test}

results <- get_estimates(10, 200, 30, iters = 200000)

mean(results$samples < .02)
```

In this case the probability of a mean prevalence below 2% is `r PooledPrevalence:::percent(mean(results$samples < .02))`, therefore there's not enough evidence in favor of removing the control measure. We may want to increase the sample size of the study or wait more. The Bayesian framework allows to simply add new samples to the already collected ones and simply update the analysis.

```{r hypothesis_test_new_data}

old.results <- get_estimates(10, 200, 30, iters = 200000)

new.tested.pools <- 60

new.positve.pools <- 9

new.results <- get_estimates(10, 200 + new.tested.pools, 30 + new.positve.pools, iters = 200000)

mean(new.results$samples < .02)
```

The new probability is `r PooledPrevalence:::percent(mean(new.results$samples < .02))`, so we have enough evidence to remove the control measure.

Finally, we want to ascertain the probability that region A, with 30 positive pools out of 200, has a lower prevalence or region B (38 positive pools out of 180).

```{r estimate_comparison}

region.A <- get_estimates(10, 200, 30, iters = 200000)

region.B <- get_estimates(10, 180, 38, iters = 200000)

mean(region.A$samples < region.B$samples)
```

## Study design optimization

The package provides a tool to help program the most efficient design for a prevalence study based on pooled testing. The methodology tries to maximize statistical accuracy while increasing the ratio between pool size and number of tests.

The user needs to specify the operational limits that are driven by the available resources (number of sampled individuals `n.max` and maximum number of laboratory test to perform `w.max`), the expected (maximal) prevalence, and by the possible reduction in sensitivity due to an excessive number of samples in the same pool (`s.max`). This number depends on a number of factors (the pathogen, the sensitivity of the test, the laboratory infrastructure and technique), therefore it needs to be decided after appropriate local validation tests.
A good starting point for validation is between 10 and 20 samples per pool.

In the following example we investigate the possible designs for a study with a maximal pool size of 15 and a maximal number of tests and individuals to sample equal to 2000.
We also set a maximal expected prevalence of 5%. Given the nature of pooled testing, the statistical accuracy decreases with the prevalence, so it is better to select a value which is higher than the prevalence we actually expect to find, in order not to end up with an underpowered study. 

```{r optimization_grid, dpi=400, fig.height=10, fig.width=10}

grid <- design_optimization(s.max = 15, w.max = 2000, p = .05, n.max = 2000)

plot_optimization_grid(grid)
```

The plot shows the "optimization windows" with different colors, ordered by the average score (lower is better).

We can observe that the best window is achieved with a pool size `s` of 10 or more and with a number of test `w` higher than 108.

Once a window has been identified, simulation tests can be used to evaluate the statistical performance. The tests need to be carried on specific couples of values of pool size `s` and number of tests `w`.

In order to decrease the risk of sensitivity problems we suggest to evaluate the more conservative designs in term of pool size (lower `s`), but validation studies can allow to choose a higher value. Regarding the number of tests, one should use the higher number that is achievable during the study, since the higher the number of tests, the higher the statistical precision; through pooling the maximum number of test needed is already decreased by a factor equal to `s` (e.g. for `s == 10`, given 2000 individuals planned for sampling, at maximum `2000 / s = 200` tests will be done).

Let's compare the results in term of statistical accuracy using `w = 109` and `w = 200`.

```{r optimization_results}

simulation.low_w <- simulate_pool_test(s = 10, w = 108, p = 0.05)

print(format_simulation(simulation.low_w))

simulation.high_w <- simulate_pool_test(s = 10, w = 200, p = 0.05)

print(format_simulation(simulation.high_w))
```

It is possible to see that in both simulations the statistical accuracy of the pooled design is similar to that achieved through an unpooled study (e.g. `s_unpooled = 1` and `w_unpooled = w * s`). Instead, it's possible to observe a definite decrease in uncertainty and error passing from 108 to 200 tests. In any case, both numbers are definitely lower than than the 2000 tests that would have been performed wihout pooling.
